---
title: "De verhouding tussen criminaliteit en vermogen per gemeente in Nederland"
author: "Roos Bakker (2890395), Lars van der Duin (2896870), Benthe tromp (2886427), Sterre de Groot (2855707), Sophie Thoolen (2863476)"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

# Set-up your environment

## Reproducibility

This project uses `renv` for package management. To restore the
environment, run in the Console:renv::restore()

```{r setup, include=FALSE}
# Packages are managed via renv.
# Run renv::restore() in the Console before knitting.
library(tidyverse)
library(readr)
library(rmarkdown)
library(yaml)
library(sf)
```

# Title Page

*Names*: Sterre de Groot, Benthe Tromp, Lars van der Duin, Sophie
Thoolen en Roos Bakker

*Group Number: 7*

*Tutorial lecturer's name* - Chantal Schouwenaar

# Part 1 - Identify a Social Problem

"*Dutch municipalities with higher income inequality have a higher
number of recorded violent crimes per 1,000 inhabitants"*

## 1.1 Describe the Social Problem

Crime and socioeconomic inequality are important social issues in the
Netherlands. Municipalities differ in terms of household wealth and the
number of registered crimes. Previous research suggests that lower
levels of wealth may be associated with higher crime rates because
economic disadvantages can lead to social exclusion, unemployment, and
fewer opportunities. Understanding this relationship is important for
policymakers and local governments, as it can help them develop more
effective strategies to reduce crime and improve social conditions. This
study investigates whether Dutch municipalities with lower household
wealth experience higher levels of registered crime per 1,000 households
between 2022 and 2024 (Nieuwbeerta et al, 2008).

# Part 2 - Data Sourcing

## 2.1 Load in the data

To reproduce, download the datasets below and save them in the `data/`
folder of the project.

| Bestand | Bron | Link |
|------------------------|------------------------|------------------------|
| misdrijven_2022_2023.csv | Data Politie | <https://data.politie.nl/#/Politie/nl/dataset/47013NED/table?ts=1780489389758> |
| Misdrijven_2024.csv | Data Politie | <https://data.politie.nl/#/Politie/nl/dataset/47013NED/table?ts=1780489389758> |
| vermogen_2022.csv | CBS StatLine | <https://opendata.cbs.nl/#/CBS/nl/dataset/86160NED/table?ts=1780405280895> |
| vermogen_2023.csv | CBS StatLine | <https://opendata.cbs.nl/#/CBS/nl/dataset/86160NED/table?ts=1780405280895> |
| Vermogen_huishoudens_2024.csv | CBS StatLine | <https://opendata.cbs.nl/#/CBS/nl/dataset/86160NED/table?ts=1780405280895> |

### *Library*

```{r load_data, include=FALSE}
library(readr)

misdrijven_2022_2023 <- read_delim("data/misdrijven_2022_2023.csv",
                          delim = ";", escape_double = FALSE,
                          trim_ws = TRUE, skip = 5)

misdrijven_2024      <- read_delim("data/Misdrijven_2024.csv",
                          delim = ";", escape_double = FALSE,
                          trim_ws = TRUE, skip = 5)

vermogen_2022        <- read_delim("data/vermogen_2022.csv",
                          delim = ";", escape_double = FALSE,
                          trim_ws = TRUE, skip = 5)

vermogen_2023        <- read_delim("data/vermogen_2023.csv",
                          delim = ";", escape_double = FALSE,
                          trim_ws = TRUE, skip = 5)

vermogen_2024        <- read_delim("data/ Vermogen_huishoudens_2024.csv",
                          delim = ";", escape_double = FALSE,
                          trim_ws = TRUE, skip = 5)

```

## 2.2 Dataset description

This project uses two public datasets. The crime data was collected by
the Dutch Police (Data Politie) and covers recorded crimes per
municipality for 2022, 2023, and 2024. The wealth data was collected by
Statistics Netherlands (CBS) and measures household wealth per
municipality for the same years. Both datasets are available at the
municipal level (*gemeente*), making them suitable for comparison across
regions.

| Dataset | Bron | Verzamelaar | Periode | Eenheid | Beperkingen |
|------------|------------|------------|------------|------------|------------|
| misdrijven | Data Politie | Politie Nederland | 2022–2024 | Gemeente | Alleen aangiftes, geen dark number |
| vermogen | CBS StatLine | CBS | 2022–2024 | Gemeente | Gebaseerd op belastingdata |

A key limitation of the crime data is that it only includes **recorded**
crimes — unregistered crimes are not captured, which may underrepresent
the true crime rate. The wealth data is derived from tax records, which
may underrepresent informal income sources.

## 2.3 Describe the type of variables included

The dataset contains both socioeconomic status (SES) information and
crime-related information. The main SES variable is household wealth,
measured by average and median household wealth per municipality. The
second key variable is the number of registered crimes per municipality.
The data does not contain individual health information. All data comes
from administrative sources collected by the Dutch Central Bureau of
Statistics (CBS) rather than interviews or surveys. The unit of
observation is the municipality, which means that each observation
represents a Dutch municipality rather than an individual person or
household.

# Part 3 - Quantifying

## 3.1 Data cleaning

The raw dataset contained missing values coded as `"."` and numeric
columns stored as text with commas as decimal separators. The following
steps were applied:

- Missing values (`"."`) were replaced with `NA`
- Column names were renamed to readable English variable names
- Numeric columns were converted from text to numeric format
- Rows with missing wealth values were removed, as these represent
  non-municipal aggregates (e.g. national totals)

### Vermogen 2022

```{r clean_vermogen_2022}
vermogen_2022_clean <- vermogen_2022 %>%
  mutate(across(everything(), ~ na_if(., "."))) %>%
  rename(
    gemeenten            = `Regio's`,
    huishoudens_x1000    = `x 1 000`,
    percentage           = `%`,
    vermogen_mld         = `mld euro`,
    vermogen_per_persoon = `1 000 euro...5`,
    vermogen_mediaan     = `1 000 euro...6`
  ) %>%
  mutate(across(c(huishoudens_x1000, vermogen_mld, vermogen_per_persoon, vermogen_mediaan),
                ~ as.numeric(str_replace(., ",", ".")))) %>%
  filter(!is.na(vermogen_mld))
 
glimpse(vermogen_2022_clean)
#%>%
#  filter(!is.na(vermogen_2022)) %>%
#  filter(vermogen_2022 != 0)
```

### Vermogen 2023

```{r clean_vermogen_2023}
vermogen_2023_clean <- vermogen_2023 %>%
  mutate(across(everything(), ~ na_if(., "."))) %>%
  rename(
    gemeenten            = `Regio's`,
    huishoudens_x1000    = `x 1 000`,
    percentage           = `%`,
    vermogen_mld         = `mld euro`,
    vermogen_per_persoon = `1 000 euro...5`,
    vermogen_mediaan     = `1 000 euro...6`
  ) %>%
  mutate(across(c(huishoudens_x1000, vermogen_mld, vermogen_per_persoon, vermogen_mediaan),
                ~ as.numeric(str_replace(., ",", ".")))) %>%
  filter(!is.na(vermogen_mld))   

glimpse(vermogen_2023_clean)
```

### Vermogen 2024

```{r clean_vermogen_2024}
vermogen_2024_clean <- vermogen_2024 %>%
  mutate(across(everything(), ~ na_if(., "."))) %>%
  rename(
    gemeenten            = `Regio's`,
    huishoudens_x1000    = `x 1 000`,
    percentage           = `%`,
    vermogen_mld         = `mld euro`,
    vermogen_per_persoon = `1 000 euro...5`,
    vermogen_mediaan     = `1 000 euro...6`
  ) %>%
  mutate(across(c(huishoudens_x1000, vermogen_mld, vermogen_per_persoon, vermogen_mediaan),
                ~ as.numeric(str_replace(., ",", ".")))) %>%
  filter(!is.na(vermogen_mld))

glimpse(vermogen_2024_clean)
```

### Misdrijven 2022 2023

```{r clean misdrijven_2022_2023}
misdrijven_2022_2023_clean <- misdrijven_2022_2023 %>% 
  mutate(across(everything(), ~ na_if(., "."))) %>% 
  rename( 
    gemeenten = `Regio's`, 
    misdrijven_2022 = `aantal...2`, 
    misdrijven_2023 = `aantal...3` 
  )
misdrijven_2022_2023_clean <- misdrijven_2022_2023_clean %>% 
  mutate(across(c(misdrijven_2022, misdrijven_2023), ~ as.numeric(str_replace(., ",", ".")))) %>% 
  filter(!is.na(misdrijven_2022))

glimpse(misdrijven_2022_2023_clean)


```

### Misdrijven 2024

```{r}
 misdrijven_2024 <- misdrijven_2024 %>%
  mutate(`Geregistreerde misdrijven` = replace(`aantal`, `aantal` == ".", NA))

misdrijven_2024_clean <- misdrijven_2024 %>%

  mutate(`aantal` = na_if(`aantal`, ".")) %>%
  rename(`Misdrijven 2024` = `aantal`) %>%
  rename(`gemeenten` = `Regio's`)

glimpse(misdrijven_2024_clean)
```

## 3.2 Generate necessary variables

### Variable 1

```{r gen_var1}
summary(vermogen_2022_clean$vermogen_per_persoon)
summary(vermogen_2023_clean$vermogen_per_persoon)
summary(vermogen_2024_clean$vermogen_per_persoon)


vermogen_2022_var1 <- vermogen_2022_clean %>%
  mutate(vermogen_groep = case_when(
    vermogen_per_persoon < 300  ~ "Low",
    vermogen_per_persoon < 500  ~ "Medium",
    TRUE                        ~ "High"
  ))

vermogen_2023_var1 <- vermogen_2023_clean %>%
  mutate(vermogen_groep = case_when(
    vermogen_per_persoon < 300  ~ "Low",
    vermogen_per_persoon < 500  ~ "Medium",
    TRUE                        ~ "High"
  ))

vermogen_2024_var1 <- vermogen_2024_clean %>%
  mutate(vermogen_groep = case_when(
    vermogen_per_persoon < 300  ~ "Low",
    vermogen_per_persoon < 500  ~ "Medium",
    TRUE                        ~ "High"
  ))
table(vermogen_2022_var1$vermogen_groep)
```

### Variable 2

Combining households from vermogen_2023 and misdrijven_2022_2023

```{r}
misdrijven_2022_2023_clean <- left_join(
  misdrijven_2022_2023,
  vermogen_2023_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2023 = huishoudens_x1000),
  by = c("Regio's" = "gemeenten")
) %>%
  left_join(
    vermogen_2022_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2022 = huishoudens_x1000),
    by = c("Regio's" = "gemeenten")
  ) %>%
  rename(
    misdrijven_2022 = aantal...2,
    misdrijven_2023 = aantal...3
  ) %>%
  mutate(
    misdrijven_per_1000_2023 = as.numeric(misdrijven_2023) / huishoudens_2023,
    misdrijven_per_1000_2022 = as.numeric(misdrijven_2022) / huishoudens_2022
  )


```

#Combining all of the data sets.

```{r}
misdrijven_2022_2024_clean <- left_join(
  misdrijven_2022_2023,
  vermogen_2023_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2023 = huishoudens_x1000),
  by = c("Regio's" = "gemeenten")
) %>%
  left_join(
    vermogen_2022_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2022 = huishoudens_x1000),
    by = c("Regio's" = "gemeenten")
  ) %>%
  left_join(
    misdrijven_2024_clean %>% select(gemeenten, `Misdrijven 2024`),
    by = c("Regio's" = "gemeenten")
  ) %>%
  left_join(
    vermogen_2024_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2024 = huishoudens_x1000),
    by = c("Regio's" = "gemeenten")
  ) %>%
  rename(
    misdrijven_2022 = aantal...2,
    misdrijven_2023 = aantal...3
  ) %>%
  mutate(
    misdrijven_per_1000_2022 = as.numeric(misdrijven_2022) / huishoudens_2022,
    misdrijven_per_1000_2023 = as.numeric(misdrijven_2023) / huishoudens_2023,
    misdrijven_per_1000_2024 = as.numeric(`Misdrijven 2024`) / huishoudens_2024
  )

```

## 3.3 Visualize temporal variation

### Line plot misdrijven

```{r}
misdrijven_lang <- misdrijven_2022_2024_clean %>%
  select(`Regio's`, starts_with("misdrijven_per_1000")) %>%
  pivot_longer(
    cols = starts_with("misdrijven_per_1000"), 
    names_to = "jaar",
    names_prefix = "misdrijven_per_1000_",
    names_transform = as.integer,
    values_to = "misdrijven_per_1000"
  ) %>%
  left_join(
    vermogen_2024_var1 %>% select(gemeenten, vermogen_groep),
    by = c("Regio's" = "gemeenten")
  ) %>%
  filter(!is.na(vermogen_groep))

misdrijven_samenvatting <- misdrijven_lang %>%
  group_by(vermogen_groep, jaar) %>%
  summarise(gem_misdrijven = mean(misdrijven_per_1000, na.rm = TRUE), .groups = "drop")

ggplot(misdrijven_samenvatting,
       aes(x = jaar, y = gem_misdrijven, colour = vermogen_groep)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = c(2022, 2023, 2024)) +
  scale_colour_manual(
    values = c(
      "Low" = "pink",
      "Medium" = "orange",
      "High" = "green"
    )
  ) +
  labs(
    title = "Crimes per 1000 households per wealth group",
    x = "Year",
    y = "Average number of crimes per 1000 households",
    colour = "Wealth group"
  ) +
  theme_minimal()

```

This plot is relevant to our social problem because it visually shows
the direct relationship between income level (low, medium, high) and
recorded crime rates in Dutch municipalities from 2022-2024. This graph
shows that the low-income group has a significantly higher number of
crimes per 1,000 inhabitants, while the middle-income and high-income
have significantly lower crime rates.

## 3.4 Visualize spatial variation

### Map misdrijven data

```{r visualise_map}
#install.packages("cbsodataR")
#install.packages("sf")

library(cbsodataR)
library(sf)
library(tidyverse)
misdrijven_gemeenten <- misdrijven_2022_2024_clean %>%
  filter(!grepl("RE|Buitenland|niet in te delen|Nederland|Bron", 
                `Regio's`, ignore.case = TRUE))
misdrijven_gemeenten <- misdrijven_gemeenten %>%
  mutate(`Regio's` = case_when(
    `Regio's` == "'s-Gravenhage (gemeente)" ~ "'s-Gravenhage",
    `Regio's` == "Groningen (gemeente)"     ~ "Groningen",
    TRUE                                    ~ `Regio's`
  ))
gemeenten_kaart <- cbs_get_sf("gemeente", 2024)
kaart_data <- gemeenten_kaart %>%
  left_join(misdrijven_gemeenten, by = c("statnaam" = "Regio's"))
ggplot(kaart_data) +
  geom_sf(aes(fill = misdrijven_per_1000_2024)) +
  scale_fill_gradient(low = "lightyellow", high = "darkred", na.value = "lightgrey") +
  labs(
    title = "Crimes per 1000 households per municipality (2024)",
    caption = "Source: CBS & Politie | Grey = no data available",
    fill = "Crimes\n(per 1000 households)"
  ) +
  theme_void()
```

This map is relevant to our social problem because it visualizes the
geographic distribution of crime rates across Dutch municipalities. This
is an important variable in our hypothesis linking income inequality to
crime. By showing where crime rates are the highest and lowest, the map
makes a visual comparison with income inequality data at the same
municipal level which will contribute to our social problem analysis.

## 3.5 Visualize sub-population variation

### Comparison misdrijven and vermogen per municipality

```{r}
ggplot(misdrijven_lang,
       aes(x = vermogen_groep, y = misdrijven_per_1000, fill = vermogen_groep)) +
  geom_boxplot() +
  facet_wrap(~ jaar) +
  scale_fill_manual(values = c(
    "Low"      = "pink",
    "Medium" = "lightyellow",
    "High"      = "lightgreen"
  )) +
  coord_cartesian(ylim = c(20,200)) +
  labs(
    title = "Crimes per 1000 households per wealth group (2022–2024)",
    x = "Wealth group",
    y = "Crimes per 1000 households",
    fill = "Wealth group"
  ) +
  theme_minimal()

```

This boxplot is relevant to our social problem because it directly
compares crime rates across municipalities grouped by wealth level (low,
medium, high). This allows us to see whether lower wealth groups are
associated with a higher crime rate compared to the medium and high
wealth groups. This supports our hypothesis that different wealth groups
are linked to a different amount of crime.

## 3.6 Event analysis
```{r}
misdrijven_lang %>%
  filter(!is.na(vermogen_groep)) %>%
  group_by(jaar, vermogen_groep) %>%
  summarise(gemiddeld = mean(misdrijven_per_1000, na.rm = TRUE), .groups = "drop") %>% 
  mutate(vermogen_groep = factor(vermogen_groep, levels = c("Low", "Medium", "High"))) %>% 
  ggplot(aes(x = jaar, y = gemiddeld, colour = vermogen_groep)) + 
  geom_vline(xintercept = 2023, linetype = "dashed", colour = "gray40", linewidth = 0.8) +
  annotate("text", x = 2023.05, y = Inf, label = "Expansion of camera surveillance", 
           hjust = 0, vjust = 1.5, size = 3, colour = "gray40") +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("Low" = "pink", "Medium" = "orange", "High" = "lightgreen")) +
  scale_x_continuous(breaks = c(2022,2023,2024)) +
  labs(
    title = "Crimes per year by wealth group",
    subtitle = "Dashed line marks the expansion of camera surveillance in 2023 ",
    x = "Year",
    y = "Average crimes per 1000 households",
    colour = "Wealth group"
  ) +
  theme_minimal()
```

This event analysis examines the impact of increased camera surveillance
in the Netherlands in 2023 on crime rates. In recent years, many
municipalities have expanded the use of surveillance cameras to improve
public safety, deter criminal activity, and support law enforcement in
identifying offenders (Koole, 2023).

The year 2023 is considered the event year, and crime figures before and
after the expansion of camera surveillance are compared. The expectation
is that increased surveillance may reduce crime by increasing the
perceived risk of being caught. However, it is also possible that
certain crimes become more frequently reported or detected due to the
presence of cameras. By analyzing crime data around this event, we can
assess whether the increase in camera surveillance is associated with a
significant change in crime levels in Dutch municipalities (Van de Ven, 2023).


# Part 4 - Discussion

## 4.1 Discuss your findings

The findings of this report suggest that municipalities with lower
household wealth generally experience higher levels of registered crime
per 1,000 households. This relationship is visible across multiple years
(2022,2023,2024) and remains relatively stable over time.

Some limitations should be acknowledged. First, the analysis only uses
registered crimes and therefore does not capture unreported crime.
Second, household wealth is only one measure of socioeconomic status and
does not fully represent inequality. Finally, this analysis covers a
period of only three years, which limits the ability to distinguish
long-term trends from short-term fluctuations.

# Part 5 - Reproducibility

## 5.1 Github repository link

Public repository: https://github.com/Sophie-Thoolen/Group-1-7---Criminaliteit-en-

## 5.2 Reference list
Flight, S. (2023, 8 augustus). Gemeentelijk cameratoezicht groeit maar door. Secondant. https://hetccv.nl/secondant-gemeentelijk-cameratoezicht-groeit-maar-door/

Van de Ven, E. (2023, 12 juni). Onderzoek 2023: Momenteel 313.354 beveiligingscamera's in Nederland. VPNGids.nl. https://www.vpngids.nl/onderzoek/onderzoek-2023-momenteel-313354-beveiligingscameras-in-nederland/

Nieuwbeerta, P., McCall, P. L., Elffers, H., & Wittebrood, K. (2008). Neighborhood characteristics and individual homicide risks: Effects of social cohesion, confidence in the police, and socioeconomic disadvantage. Homicide Studies, 12(1), 90–116. https://doi.org/10.1177/1088767907310913
