---
title: "De verhouding tussen criminaliteit en vermogen per gemeente in Nederland"
author: "Roos Bakker (2890395), Lars van der Duin (2896870), Benthe tromp (2886427), Sterre de Groot (2855707), Sophie Thoolen (2863476)"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Title Page

*Names*: Sterre de Groot, Benthe Tromp, Lars van der Duin, Sophie
Thoolen en Roos Bakker

*Group Number: 7*

*Tutorial lecturer's name* - Chantal Schouwenaar

# Part 1 - Identify a Social Problem

"*Dutch municipalities with higher income inequality have a higher
number of recorded violent crimes per 1,000 inhabitants"*

Use APA referencing throughout your document. [Here's a link to some
explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

- Why is this relevant?

- ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

To reproduce download the data underneath and save them in the 'data/'
map.

misdrijven_2022_2023.csv \| Data Politie \|
<https://data.politie.nl/#/Politie/nl/dataset/47013NED/table?ts=1780489389758>
\|

Misdrijven_2024.csv \| Data Politie\|
<https://data.politie.nl/#/Politie/nl/dataset/47013NED/table?ts=1780489389758>
\| vermogen_2022.csv \|

CBS StatLine \|
<https://opendata.cbs.nl/#/CBS/nl/dataset/86160NED/table?ts=1780405280895>
\| vermogen_2023.csv \|

CBS StatLine \|
<https://opendata.cbs.nl/#/CBS/nl/dataset/86160NED/table?ts=1780405280895>\|
Vermogen_huishoudens_2024.csv \|

CBS StatLine \|
<https://opendata.cbs.nl/#/CBS/nl/dataset/86160NED/table?ts=1780405280895>\|

### *Library*

```{r load data}
library(readr)
Misdrijven_2024 <- read_delim("data/Misdrijven_2024.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE,skip = 3)
misdrijven_2024 <- read_delim("data/misdrijven_2024.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE,skip = 3)
 
library(readr)
misdrijven_2022_2023 <- read_delim("data/misdrijven_2022_2023.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 3)

library(readr)
X_Vermogen_huishoudens_2024 <- read_delim("data/ Vermogen_huishoudens_2024.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)

library(readr)
vermogen_2023 <- read_delim("data/vermogen_2023.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)

library(readr)
vermogen_2022 <- read_delim("data/vermogen_2022.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)

```

In this case we see 28 variables, but we miss some information on what
units they are in. We also don't know anything about the year/moment in
which this data has been captured.

These are things that are usually included in the metadata of the
dataset. For your project, you need to provide us with the information
from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data
  coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

### Cleaned version vermogen 2022

```{r}
library(readr)
vermogen_2022 <- read_delim("data/vermogen_2022.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)
#View(vermogen_2022) 
install.packages("tidyverse")
library(tidyverse)

vermogen_2022_clean <- vermogen_2022 %>%
  mutate(across(everything(), ~ na_if(., "."))) %>%
  rename(
    gemeenten            = `Regio's`,
    huishoudens_x1000    = `x 1 000`,
    percentage           = `%`,
    vermogen_mld         = `mld euro`,
    vermogen_per_persoon = `1 000 euro...5`,
    vermogen_mediaan     = `1 000 euro...6`
  ) %>%
  mutate(across(c(huishoudens_x1000, vermogen_mld, vermogen_per_persoon, vermogen_mediaan),
                ~ as.numeric(str_replace(., ",", ".")))) %>%
  filter(!is.na(vermogen_mld))
 
glimpse(vermogen_2022_clean)

View(vermogen_2022_clean)

colnames(vermogen_2022)
colnames(vermogen_2022_clean)

#%>%
#  filter(!is.na(vermogen_2022)) %>%
#  filter(vermogen_2022 != 0)
```

### Cleaned version vermogen 2023

```{r}
library(readr)
library(tidyverse)

vermogen_2023 <- read_delim("data/vermogen_2023.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)

vermogen_2023_clean <- vermogen_2023 %>%
  mutate(across(everything(), ~ na_if(., "."))) %>%
  rename(
    gemeenten            = `Regio's`,
    huishoudens_x1000    = `x 1 000`,
    percentage           = `%`,
    vermogen_mld         = `mld euro`,
    vermogen_per_persoon = `1 000 euro...5`,
    vermogen_mediaan     = `1 000 euro...6`
  ) %>%
  mutate(across(c(huishoudens_x1000, vermogen_mld, vermogen_per_persoon, vermogen_mediaan),
                ~ as.numeric(str_replace(., ",", ".")))) %>%
  filter(!is.na(vermogen_mld))    # ← kleine letter v, gefixt!

glimpse(vermogen_2023_clean)
View(vermogen_2023_clean)
```

### Cleaned version vermogen 2024

```{r}
library(readr)
library(tidyverse)

vermogen_2024 <- read_delim("data/ Vermogen_huishoudens_2024.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)

vermogen_2024_clean <- vermogen_2024 %>%
  mutate(across(everything(), ~ na_if(., "."))) %>%
  rename(
    gemeenten            = `Regio's`,
    huishoudens_x1000    = `x 1 000`,
    percentage           = `%`,
    vermogen_mld         = `mld euro`,
    vermogen_per_persoon = `1 000 euro...5`,
    vermogen_mediaan     = `1 000 euro...6`
  ) %>%
  mutate(across(c(huishoudens_x1000, vermogen_mld, vermogen_per_persoon, vermogen_mediaan),
                ~ as.numeric(str_replace(., ",", ".")))) %>%
  filter(!is.na(vermogen_mld))

glimpse(vermogen_2024_clean)
View(vermogen_2024_clean)
```

### Cleaned version 2022 2023 vermogen

```{r}
library(readr)
library(tidyverse)

misdrijven_2022_2023 <- read_delim("data/misdrijven_2022_2023.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)

misdrijven_2022_2023_clean <- misdrijven_2022_2023 %>%
  mutate(across(everything(), ~ na_if(., "."))) %>%
  rename(
    gemeenten       = `Regio's`,
    misdrijven_2022 = `aantal...2`,
    misdrijven_2023 = `aantal...3`
  )

misdrijven_2022_2023_clean <- misdrijven_2022_2023_clean %>%
  mutate(across(c(misdrijven_2022, misdrijven_2023),
                ~ as.numeric(str_replace(., ",", ".")))) %>%
  filter(!is.na(misdrijven_2022))    # ← kleine letter m!

glimpse(misdrijven_2022_2023_clean)
View(misdrijven_2022_2023_clean)
```

### Cleaned version 2024 vermogen

```{r}
library(readr)
Misdrijven_2024 <- read_delim("data/Misdrijven_2024.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE,skip = 3)
View(Misdrijven_2024)
misdrijven_2024 <- read_delim("data/misdrijven_2024.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE,skip = 3)
View(misdrijven_2024)

install.packages("tidyverse")
library(tidyverse)

misdrijven_2024 <- misdrijven_2024 %>%
  mutate(`Geregistreerde misdrijven` = replace(`Geregistreerde misdrijven`, `Geregistreerde misdrijven` == ".", NA))

misdrijven_2024_clean <- misdrijven_2024 %>%
  mutate(`Geregistreerde misdrijven` = na_if(`Geregistreerde misdrijven`, ".")) %>%
  rename(`Misdrijven 2024` = `Geregistreerde misdrijven`) %>%
  rename(`gemeenten` = `Onderwerp`)

View (Misdrijven_2024_clean)
View (misdrijven_2024_clean)
```

## 3.2 Generate necessary variables

### Variable 1

```{r gen_var1}
summary(vermogen_2022_clean$vermogen_per_persoon)
summary(vermogen_2023_clean$vermogen_per_persoon)
summary(vermogen_2024_clean$vermogen_per_persoon)


vermogen_2022_clean <- vermogen_2022_clean %>%
  mutate(vermogen_groep = case_when(
    vermogen_per_persoon < 300  ~ "Laag",
    vermogen_per_persoon < 500  ~ "Gemiddeld",
    TRUE                        ~ "Hoog"
  ))

vermogen_2023_clean <- vermogen_2023_clean %>%
  rename_with(tolower)

vermogen_2023_clean <- vermogen_2023_clean %>%
  mutate(vermogen_groep = case_when(
    vermogen_per_persoon < 300  ~ "Laag",
    vermogen_per_persoon < 500  ~ "Gemiddeld",
    TRUE                        ~ "Hoog"
  ))

vermogen_2024_clean <- vermogen_2024_clean %>%
  mutate(vermogen_groep = case_when(
    vermogen_per_persoon < 300  ~ "Laag",
    vermogen_per_persoon < 500  ~ "Gemiddeld",
    TRUE                        ~ "Hoog"
  ))
table(vermogen_2022_clean$vermogen_groep)
View(vermogen_2022_clean)
```

### Variable 2

Combining huishoudens from vermogen_2023 (cleaned version) and
misdrijven_2022_202

```{r}
misdrijven_2022_2023_clean <- left_join(
  misdrijven_2022_2023,
  vermogen_2023_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2023 = huishoudens_x1000),
  by = c("Regio's" = "gemeenten")
) %>%
  left_join(
    vermogen_2022_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2022 = huishoudens_x1000),
    by = c("Regio's" = "gemeenten")
  ) %>%
  rename(
    misdrijven_2022 = aantal...2,
    misdrijven_2023 = aantal...3
  ) %>%
  mutate(
    misdrijven_per_1000_2023 = as.numeric(misdrijven_2023) / huishoudens_2023,
    misdrijven_per_1000_2022 = as.numeric(misdrijven_2022) / huishoudens_2022
  )


```

```{r}
misdrijven_2022_2024_clean <- left_join(
  misdrijven_2022_2023,
  vermogen_2023_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2023 = huishoudens_x1000),
  by = c("Regio's" = "gemeenten")
) %>%
  left_join(
    vermogen_2022_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2022 = huishoudens_x1000),
    by = c("Regio's" = "gemeenten")
  ) %>%
  left_join(
    misdrijven_2024_clean %>% select(gemeenten, `Misdrijven 2024`),
    by = c("Regio's" = "gemeenten")
  ) %>%
  left_join(
    vermogen_2024_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2024 = huishoudens_x1000),
    by = c("Regio's" = "gemeenten")
  ) %>%
  rename(
    misdrijven_2022 = aantal...2,
    misdrijven_2023 = aantal...3
  ) %>%
  mutate(
    misdrijven_per_1000_2022 = as.numeric(misdrijven_2022) / huishoudens_2022,
    misdrijven_per_1000_2023 = as.numeric(misdrijven_2023) / huishoudens_2023,
    misdrijven_per_1000_2024 = as.numeric(`Misdrijven 2024`) / huishoudens_2024
  )

```

## 3.3 Visualize temporal variation

### Line plot misdrijven

```{r}
misdrijven_lang %>%
  left_join(df_merged %>% select(gemeenten, vermogen_groep), 
            by = c("Regio's" = "gemeenten")) %>%
  filter(!is.na(vermogen_groep)) %>%
  group_by(jaar, vermogen_groep) %>%
  summarise(gemiddeld = mean(misdrijven_per_1000, na.rm = TRUE)) %>%
  mutate(vermogen_groep = factor(vermogen_groep, levels = c("Laag", "Gemiddeld", "Hoog"))) %>%
  ggplot(aes(x = jaar, y = gemiddeld, colour = vermogen_groep)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("Laag" = "lightpink", "Gemiddeld" = "orange", "Hoog" = "lightgreen")) +
  scale_x_continuous(breaks = c(2022, 2023, 2024)) +
  labs(
    title = "Misdrijven per jaar per vermogensgroep",
    x = "Jaar",
    y = "Gemiddelde misdrijven per 1000 huishoudens",
    colour = "Vermogensgroep"
  ) +
  theme_minimal()

```

## 3.4 Visualize spatial variation

### Map misdrijven data

```{r visualise_map}
#install.packages("cbsodataR")
#install.packages("sf")

library(cbsodataR)
library(sf)
library(tidyverse)
misdrijven_gemeenten <- misdrijven_2022_2024_clean %>%
  filter(!grepl("RE|Buitenland|niet in te delen|Nederland|Bron", 
                `Regio's`, ignore.case = TRUE))
misdrijven_gemeenten <- misdrijven_gemeenten %>%
  mutate(`Regio's` = case_when(
    `Regio's` == "'s-Gravenhage (gemeente)" ~ "'s-Gravenhage",
    `Regio's` == "Groningen (gemeente)"     ~ "Groningen",
    TRUE                                    ~ `Regio's`
  ))
kaart_data <- gemeenten_kaart %>%
  left_join(misdrijven_gemeenten, by = c("statnaam" = "Regio's"))
ggplot(kaart_data) +
  geom_sf(aes(fill = misdrijven_per_1000_2024)) +
  scale_fill_gradient(low = "lightyellow", high = "darkred", na.value = "lightgrey") +
  labs(
    title = "Misdrijven per 1000 huishoudens per gemeente (2024)",
    caption = "Bron: CBS & Politie | Grijs = geen data beschikbaar",
    fill = "Misdrijven\n(per 1000 huishoudens)"
  ) +
  theme_void()
```

### Map vermogen data

Here you provide a description of why the plot above is relevant to your
specific social problem.

```{r}
library(tidyverse)
library(sf)
library(cbsodataR)
gemeenten_kaart <- cbs_get_sf("gemeente", 2024)
vermogen_2022_gem <- vermogen_2022_clean %>%
  filter(!grepl("LD|Buitenland|Nederland|Bron", gemeenten, ignore.case = TRUE)) %>%
  select(gemeenten, vermogen_per_persoon) %>%
  rename(vermogen_2022 = vermogen_per_persoon)

vermogen_2023_gem <- vermogen_2023_clean %>%
  filter(!grepl("LD|Buitenland|Nederland|Bron", gemeenten, ignore.case = TRUE)) %>%
  select(gemeenten, vermogen_per_persoon) %>%
  rename(vermogen_2023 = vermogen_per_persoon)

vermogen_2024_gem <- vermogen_2024_clean %>%
  filter(!grepl("LD|Buitenland|Nederland|Bron", gemeenten, ignore.case = TRUE)) %>%
  select(gemeenten, vermogen_per_persoon) %>%
  rename(vermogen_2024 = vermogen_per_persoon)

vermogen_gemiddeld <- vermogen_2022_gem %>%
  left_join(vermogen_2023_gem, by = "gemeenten") %>%
  left_join(vermogen_2024_gem, by = "gemeenten") %>%
  mutate(vermogen_gemiddeld = (vermogen_2022 + vermogen_2023 + vermogen_2024) / 3)

vermogen_gemiddeld <- vermogen_gemiddeld %>%
  mutate(gemeenten = case_when(
    gemeenten == "'s-Gravenhage (gemeente)" ~ "Den Haag",
    gemeenten == "Groningen (gemeente)"     ~ "Groningen",
    gemeenten == "Utrecht (gemeente)"       ~ "Utrecht",
    TRUE                                    ~ gemeenten
  ))
kaart_vermogen <- gemeenten_kaart %>%
  left_join(vermogen_gemiddeld, by = c("statnaam" = "gemeenten"))


ggplot(kaart_vermogen) +
  geom_sf(aes(fill = vermogen_gemiddeld)) +
scale_fill_gradient(low = "lightyellow", high = "darkblue", na.value = "lightgrey") +
  labs(
    title = "Gemiddeld vermogen per huishouden per gemeente (2024)",
    caption = "Bron: CBS | Grijs = geen data beschikbaar",
    fill = "Vermogen\n(x 1000 euro)"
  ) +
  theme_void()  
```

Here you provide a description of why the plot above is relevant to your
specific social problem.

## 3.5 Visualize sub-population variation

### Comparison misdrijven and vermogen per municipality

```{r}
df_merged <- vermogen_gemiddeld %>%
  left_join(
    misdrijven_gemeenten %>% select(`Regio's`, misdrijven_per_1000_2024),
    by = c("gemeenten" = "Regio's")
  ) %>%
  left_join(
    vermogen_2024_clean %>% select(gemeenten, vermogen_groep),
    by = "gemeenten"
  )
df_merged <- df_merged %>%
  mutate(vermogen_groep = factor(vermogen_groep, levels = c("Laag", "Gemiddeld", "Hoog")))
ggplot(df_merged, aes(x = vermogen_groep, y = misdrijven_per_1000_2024, fill = vermogen_groep)) +
  geom_boxplot(colour = "darkblue", width = 0.5) +
  scale_fill_manual(values = c("Laag" = "lightpink", "Gemiddeld" = "lightyellow", "Hoog" = "lightgreen")) +
  labs(
    title = "Misdrijven per vermogensgroep (2024)",
    x = "Vermogensgroep",
    y = "Misdrijven per 1000 huishoudens",
    fill = "Vermogensgroep"
  ) +
  theme_minimal()
  
```

Here you provide a description of why the plot above is relevant to your
specific social problem.

## 3.6 Event analysis

bron 1:
<https://www.vpngids.nl/onderzoek/onderzoek-2023-momenteel-313354-beveiligingscameras-in-nederland/>
bron 2:
<https://hetccv.nl/secondant-gemeentelijk-cameratoezicht-groeit-maar-door/>

```{r}
misdrijven_lang %>% 
  left_join(df_merged %>% select(gemeenten, vermogen_groep), by = c("Regio's" = "gemeenten")) %>% 
  filter(!is.na(vermogen_groep)) %>% 
  group_by(jaar, vermogen_groep) %>% 
  summarise(gemiddeld = mean(misdrijven_per_1000, na.rm = TRUE), .groups = "drop") %>% 
  mutate(vermogen_groep = factor(vermogen_groep, levels = c("Laag", "Gemiddeld", "Hoog"))) %>% 
  ggplot(aes(x = jaar, y = gemiddeld, colour = vermogen_groep)) + 
  geom_vline(xintercept = 2023, linetype = "dashed", colour = "gray40", linewidth = 0.8) +  # verticale lijn
  annotate("text", x = 2023.05, y = Inf, label = "Uitbreiding cameratoezicht", 
           hjust = 0, vjust = 1.5, size = 3, colour = "gray40") + 
  geom_line(linewidth = 1.2) + 
  geom_point(size = 3) + 
  scale_colour_manual(values = c("Laag" = "lightpink", "Gemiddeld" = "orange", "Hoog" = "lightgreen")) + 
  scale_x_continuous(breaks = c(2022, 2023, 2024)) + 
  labs(
    title = "Misdrijven per jaar per vermogensgroep",
    subtitle = "Verticale lijn markeert uitbreiding cameratoezicht in 2023",
    x = "Jaar", 
    y = "Gemiddelde misdrijven per 1000 huishoudens", 
    colour = "Vermogensgroep"
  ) + 
  theme_minimal()
```

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your
specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

The findings of this report suggest that municipalities with lower
household wealth generally experience higher levels of registered crime
per 1,000 households. This relationship is visible across multiple years
(2022,2023,2024) and remains relatively stable over time.

Some limitations should be acknowledged. First, the analysis only uses
registered crimes and therefore does not capture unreported crime.
Second, household wealth is only one measure of socioeconomic status and
does not fully represent inequality. Finally, this analysis covers a
period of only three years, which limits the ability to distinguish
long-term trends from short-term fluctuations.

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
