---
title: "De verhouding tussen criminaliteit en vermogen per gemeente in Nederland"
author: "Roos Bakker (2890395), Lars van der Duin (2896870), Benthe tromp (2886427), Sterre de Groot (2855707), Sophie Thoolen (2863476)"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Title Page

Include your names
Sterre de Groot, Benthe Tromp, Lars van der Duin, Sophie Thoolen en Roos Bakker 
Include the tutorial group 
Number 7

Include your tutorial lecturer's name
- Chantal Schouwenaar

# Part 1 - Identify a Social Problem
"Dutch municipalities with higher income inequality have a higher number of recorded violent crimes per 1,000 inhabitants"

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

###datasets
```{r}
library(readr)
Misdrijven_2024 <- read_delim("data/Misdrijven_2024.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE,skip = 3)
View(Misdrijven_2024)
misdrijven_2024 <- read_delim("data/misdrijven_2024.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE,skip = 3)
View(misdrijven_2024)
 
library(readr)
misdrijven_2022_2023 <- read_delim("data/misdrijven_2022_2023.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 3)
View(misdrijven_2022_2023)

library(readr)
X_Vermogen_huishoudens_2024 <- read_delim("data/ Vermogen_huishoudens_2024.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)
View(X_Vermogen_huishoudens_2024)

library(readr)
vermogen_2023 <- read_delim("data/vermogen_2023.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)
View(vermogen_2023)

library(readr)
vermogen_2022 <- read_delim("data/vermogen_2022.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)
View(vermogen_2022) 

```

#Gecleande versies 
##2022 Vermogen clean
```{r}
library(readr)
vermogen_2022 <- read_delim("data/vermogen_2022.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)
#View(vermogen_2022) 
install.packages("tidyverse")
library(tidyverse)

vermogen_2022_clean <- vermogen_2022 %>%
  mutate(across(everything(), ~ na_if(., "."))) %>%
  rename(
    gemeenten            = `Regio's`,
    huishoudens_x1000    = `x 1 000`,
    percentage           = `%`,
    vermogen_mld         = `mld euro`,
    vermogen_per_persoon = `1 000 euro...5`,
    vermogen_mediaan     = `1 000 euro...6`
  ) %>%
  mutate(across(c(huishoudens_x1000, vermogen_mld, vermogen_per_persoon, vermogen_mediaan),
                ~ as.numeric(str_replace(., ",", ".")))) %>%
  filter(!is.na(vermogen_mld))
 
glimpse(vermogen_2022_clean)

View(vermogen_2022_clean)

colnames(vermogen_2022)
colnames(vermogen_2022_clean)

#%>%
#  filter(!is.na(vermogen_2022)) %>%
#  filter(vermogen_2022 != 0)
```

## 2023 Vermogen clean
```{r}
library(readr)
library(tidyverse)

vermogen_2023 <- read_delim("data/vermogen_2023.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)

vermogen_2023_clean <- vermogen_2023 %>%
  mutate(across(everything(), ~ na_if(., "."))) %>%
  rename(
    gemeenten            = `Regio's`,
    huishoudens_x1000    = `x 1 000`,
    percentage           = `%`,
    vermogen_mld         = `mld euro`,
    vermogen_per_persoon = `1 000 euro...5`,
    vermogen_mediaan     = `1 000 euro...6`
  ) %>%
  mutate(across(c(huishoudens_x1000, vermogen_mld, vermogen_per_persoon, vermogen_mediaan),
                ~ as.numeric(str_replace(., ",", ".")))) %>%
  filter(!is.na(vermogen_mld))    # ← kleine letter v, gefixt!

glimpse(vermogen_2023_clean)
View(vermogen_2023_clean)
```

##2024 Vermogen clean
```{r}
library(readr)
library(tidyverse)

vermogen_2024 <- read_delim("data/ Vermogen_huishoudens_2024.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)

vermogen_2024_clean <- vermogen_2024 %>%
  mutate(across(everything(), ~ na_if(., "."))) %>%
  rename(
    gemeenten            = `Regio's`,
    huishoudens_x1000    = `x 1 000`,
    percentage           = `%`,
    vermogen_mld         = `mld euro`,
    vermogen_per_persoon = `1 000 euro...5`,
    vermogen_mediaan     = `1 000 euro...6`
  ) %>%
  mutate(across(c(huishoudens_x1000, vermogen_mld, vermogen_per_persoon, vermogen_mediaan),
                ~ as.numeric(str_replace(., ",", ".")))) %>%
  filter(!is.na(vermogen_mld))

glimpse(vermogen_2024_clean)
View(vermogen_2024_clean)
```

## 2022_2023 Misdrijven clean
```{r}
library(readr)
library(tidyverse)

misdrijven_2022_2023 <- read_delim("data/misdrijven_2022_2023.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
    skip = 5)

misdrijven_2022_2023_clean <- misdrijven_2022_2023 %>%
  mutate(across(everything(), ~ na_if(., "."))) %>%
  rename(
    gemeenten       = `Regio's`,
    misdrijven_2022 = `aantal...2`,
    misdrijven_2023 = `aantal...3`
  )

misdrijven_2022_2023_clean <- misdrijven_2022_2023_clean %>%
  mutate(across(c(misdrijven_2022, misdrijven_2023),
                ~ as.numeric(str_replace(., ",", ".")))) %>%
  filter(!is.na(misdrijven_2022))    # ← kleine letter m!

glimpse(misdrijven_2022_2023_clean)
View(misdrijven_2022_2023_clean)
```

##2024 Misdrijven clean
```{r}
library(readr)
Misdrijven_2024 <- read_delim("data/Misdrijven_2024.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE,skip = 3)
View(Misdrijven_2024)
misdrijven_2024 <- read_delim("data/misdrijven_2024.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE,skip = 3)
View(misdrijven_2024)

install.packages("tidyverse")
library(tidyverse)

Misdrijven_2024 <- Misdrijven_2024 %>%
misdrijven_2024 <- misdrijven_2024 %>%
  mutate(`Geregistreerde misdrijven` = replace(`Geregistreerde misdrijven`, `Geregistreerde misdrijven` == ".", NA))

Misdrijven_2024_clean <- Misdrijven_2024 %>%
misdrijven_2024_clean <- misdrijven_2024 %>%
  mutate(`Geregistreerde misdrijven` = na_if(`Geregistreerde misdrijven`, ".")) %>%
  rename(`Misdrijven 2024` = `Geregistreerde misdrijven`) %>%
  rename(`Gemeenten` = `Onderwerp`)

View (Misdrijven_2024_clean)
View (misdrijven_2024_clean)
```

```
```

```{r}


#midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)



```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

#Variable 1
```{r gen_var1}
summary(vermogen_2022_clean$vermogen_per_persoon)
summary(vermogen_2023_clean$vermogen_per_persoon)
summary(vermogen_2024_clean$vermogen_per_persoon)


vermogen_2022_clean <- vermogen_2022_clean %>%
  mutate(vermogen_groep = case_when(
    vermogen_per_persoon < 300  ~ "Laag",
    vermogen_per_persoon < 500  ~ "Gemiddeld",
    TRUE                        ~ "Hoog"
  ))

vermogen_2023_clean <- vermogen_2023_clean %>%
  rename_with(tolower)

vermogen_2023_clean <- vermogen_2023_clean %>%
  mutate(vermogen_groep = case_when(
    vermogen_per_persoon < 300  ~ "Laag",
    vermogen_per_persoon < 500  ~ "Gemiddeld",
    TRUE                        ~ "Hoog"
  ))

vermogen_2024_clean <- vermogen_2024_clean %>%
  mutate(vermogen_groep = case_when(
    vermogen_per_persoon < 300  ~ "Laag",
    vermogen_per_persoon < 500  ~ "Gemiddeld",
    TRUE                        ~ "Hoog"
  ))
table(vermogen_2022_clean$vermogen_groep)
View(vermogen_2022_clean)
```

#huishoudens combineren met misdrijven 2024
```{r}
vermogen_2024_clean <- vermogen_2024_clean  %>% select(gemeenten, huishoudens_x1000)
huishoudens_misdrijven_2024 <- left_join(
<<<<<<< HEAD
  Misdrijven_2024_clean,
  vermogen_2024_clean %>% select(gemeenten, huishoudens_x1000),
  by = "gemeenten"
=======
  Misdrijven_2024_clean,
  vermogen_2024_clean,
  misdrijven_2024_clean,
  vermogen_2024_clean %>% select(Gemeenten, Huishoudens_x1000),
  by = "Gemeenten"
>>>>>>> a417d03361ed345a59101a10556ee3556e181cd1
)
Huishoudens_misdrijven_2024 <- Huishoudens_misdrijven_2024 %>%
huishoudens_misdrijven_2024 <- Huishoudens_misdrijven_2024 %>%
  mutate(
    `Misdrijven 2024` = as.numeric(`Misdrijven 2024`),
    Huishoudens_x1000 = as.numeric(Huishoudens_x1000),
    Misdrijven_per_huishouden = `Misdrijven 2024` / (Huishoudens_x1000 * 1000)
  )
View(Huishoudens_misdrijven_2024)
View(huishoudens_misdrijven_2024)
```


#huishoudens combineren in misdrijven 2022 en 2023
```{r}
misdrijven_2022_2023_clean <- left_join(
  misdrijven_2022_2023,
  vermogen_2023_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2023 = huishoudens_x1000),
  by = c("Regio's" = "Gemeenten")
) %>%
  left_join(
    vermogen_2022_clean %>% select(gemeenten, huishoudens_x1000) %>% rename(huishoudens_2022 = huishoudens_x1000),
    by = c("regio's" = "gemeenten")
  )

View(misdrijven_2022_2023_clean)
```


#Variable 2 
#misdrijven per huishouden 2024
```{r gen_var2}
gecombineerd <- gecombineerd %>%
  mutate(misdrijven_per_1000_huishoudens = as.numeric(`Misdrijven 2024`) / Huishoudens_x1000)

```
#misdrijven per huishouden 2022_2023
```{r}
misdrijven_2022_2023_clean <- misdrijven_2022_2023_clean %>%
  mutate(
    misdrijven_per_1000_huishoudens_2022 = as.numeric(aantal...2) / Huishoudens_2022,
    misdrijven_per_1000_huishoudens_2023 = as.numeric(aantal...3) / Huishoudens_2023
  )

View(misdrijven_2022_2023_clean)
```


## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
